A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-22 paper.Rmd
done with susans suggestions for introduction

Diagnostics are the key to determining whether there is anything importantly wrong with a model. In linear regression analysis, studying the residuals from a model fit is a common diagnostic activity. Residuals summarise what is not captured by the model, and thus provide the capacity to identify what might be wrong.

We can assess residuals in multiple ways. To examine the their univariate distribution, residuals may be plotted as a histogram or normal probability plot. Using the classical normal linear regression model as an example, if the distribution is symmetric and unimodal, we would consider it to be well-behaved. However, if the distribution is skewed, bimodal, multimodal, or contains outliers, there would be cause for concern. One could also inspect the distribution by conducting a goodness-of-fit test, such as the Shapiro-Wilk (SW) Normality test [@shapiro1965analysis].

In addition, scatterplots of residuals plotted against the fitted values and each of the explanatory variables, are made to scrutinize their relationships. If there are any visually discoverable patterns, the model is potentially inadequate or incorrectly specified. In general, one looks for noticeable departures from the model such as non-linear dependency or heteroskedasticity. A non-linear dependency would suggest that the model needs to have some additional non-linear terms. Heteroskedasticity suggests that the error is dependent on the predictors, and hence violates the independence assumption. However, correctly judging whether NO pattern exists in a residual plot is a difficult task for humans. We humans will almost always will see a pattern, so the question that really needs answering is whether any pattern perceived is consistent with randomness, purely sampling variability or noise. It is especially difficult to teach this to new analysts and students [@loy2021bringing]. To answer this, there have been numerous conventional hypothesis tests made available to test for non-linear dependence [e.g. @ramsey_tests_1969], and heteroskedasticity  [e.g. @breusch_simple_1979].

Linear regression is a well-established procedure, and there is considerable literature describing diagnostic procedures, e.g. @draper1998applied, @montgomery1982introduction, @belsley_regression_1980, @cook_applied_1999 and @cook1982residuals. An interesting details, is that despite the abundance of conventional tests, ALL of these writings advise that plotting residuals is an essential tool for diagnosing regression model problems. @draper1998applied and @belsley_regression_1980 explain that residual plots are usually revealing when the assumptions are violated. @cook_applied_1999 thinks formal tests and graphical procedures are complementary and both have a place in residual analysis, but they focus on graphical methods rather than on formal testing. @montgomery1982introduction even suggests that residual plots are more informative in most practical situations than the corresponding conventional hypothesis tests.

The common wisdom of experts is that the optimal method for diagnosing model fits is by plotting the data. The persistence of this advice to check the plots is *curious*, and investigating why this is, is the subject of this paper.

@@

Graphical summaries where residuals are plotted against fitted values, or other functions of the predictors (expected to be approximately orthogonal to the residuals) are referred to as standard residual plots in @cook1982residuals. Figure \ref{fig:residual-plot-common-departures}A shows an example of an ideal residual plot where points are symmetrically distributed around the horizontal zero line (red), with no discernible patterns. There can be various types of departures from this ideal pattern. Non-linearity, heteroskedasticity and non-normality are three commonly checked departures.

>>>>

Diagnosing a model is an important part of analysing data. In linear regression analysis, studying the residuals from a model fit is a common diagnostic activity. Residuals summarise what is not captured by the model, and thus provide the capacity to identify what might be wrong.

We can assess residuals in multiple ways. To examine the their univariate distribution, residuals may be plotted as a histogram or normal probability plot. Using the classical normal linear regression model as an example, if the distribution is symmetric and unimodal, we would consider it to be well-behaved. However, if the distribution is skewed, bimodal, multimodal, or contains outliers, there would be cause for concern. We can also inspect the distribution by conducting a goodness-of-fit test, such as the Shapiro-Wilk (SW) normality test [@shapiro1965analysis].

In addition, scatterplots of residuals plotted against the fitted values and each of the explanatory variables, can be used to scrutinize their relationships. If there are any visually discoverable patterns, the model is potentially inadequate or incorrectly specified. In general, one looks for noticeable departures from the model such as non-linear pattern or heteroskedasticity. A non-linear pattern would suggest that the model needs to have some additional non-linear terms. Heteroskedasticity suggests that the error is dependent on the predictors, and hence violates the independence assumption. However, correctly judging whether NO pattern exists in a residual plot is a difficult task for humans. Humans will almost always will see a pattern (see @kahneman), so the question that really needs answering is whether any pattern perceived is consistent with randomness, purely sampling variability, or noise. It is especially difficult to teach this to new analysts and students [@loy2021bringing]. To address this, statistical tests are available to check for non-linear patterns [e.g. @ramsey_tests_1969], and heteroskedasticity  [e.g. @breusch_simple_1979].

Linear regression is a well-established procedure, and there is considerable literature describing diagnostic procedures, e.g. @draper1998applied, @montgomery1982introduction, @belsley_regression_1980, @cook_applied_1999 and @cook1982residuals. Interestingly, despite the abundance of conventional tests, ALL of these writings advise that plotting residuals is an essential tool for diagnosing regression model problems:

> *Some most useful checks allowed by data plots should be done on a routine basis for every regression.* [@draper1998applied]

> *Such formal and informal procedures are complementary, and both have a place in residual analysis.* [@cook1982residuals]

> *In our experience, statistical tests on regression model residuals are not widely used. In most practical situations the residual plots are more informative than the corresponding tests. However, since residual plots do require skill and experience to interpret, the statistical tests may occasionally prove useful.* [@montgomery1982introduction]

\noindent The common wisdom of experts is that plotting the residuals is indispensable for diagnosing model fits. The ubiquity of this advice is *curious*: investigating why, is the subject of this paper.

@@

Graphical summaries where residuals are plotted against fitted values, or other functions of the predictors (expected to be approximately orthogonal to the residuals) are considered to be the most important residual plots by @cook_applied_1999. Figure \ref{fig:residual-plot-common-departures}A shows an example of an ideal residual plot where points are symmetrically distributed around the horizontal zero line (red), with no discernible patterns. There can be various types of departures from this ideal pattern. Non-linearity, heteroskedasticity and non-normality are three commonly checked departures.

### 2023-04-22 paper.tex
done with susans suggestions for introduction

Diagnostics are the key to determining whether there is anything
importantly wrong with a model. In linear regression analysis, studying
the residuals from a model fit is a common diagnostic activity.
Residuals summarise what is not captured by the model, and thus provide
the capacity to identify what might be wrong.

We can assess residuals in multiple ways. To examine the their
univariate distribution, residuals may be plotted as a histogram or
normal probability plot. Using the classical normal linear regression
model as an example, if the distribution is symmetric and unimodal, we
would consider it to be well-behaved. However, if the distribution is
skewed, bimodal, multimodal, or contains outliers, there would be cause
for concern. One could also inspect the distribution by conducting a
goodness-of-fit test, such as the Shapiro-Wilk (SW) Normality test
\citep{shapiro1965analysis}.

In addition, scatterplots of residuals plotted against the fitted values
and each of the explanatory variables, are made to scrutinize their
relationships. If there are any visually discoverable patterns, the
model is potentially inadequate or incorrectly specified. In general,
one looks for noticeable departures from the model such as non-linear
dependency or heteroskedasticity. A non-linear dependency would suggest
that the model needs to have some additional non-linear terms.
Heteroskedasticity suggests that the error is dependent on the
predictors, and hence violates the independence assumption. However,
correctly judging whether NO pattern exists in a residual plot is a
difficult task for humans. We humans will almost always will see a
pattern, so the question that really needs answering is whether any
pattern perceived is consistent with randomness, purely sampling
variability or noise. It is especially difficult to teach this to new
analysts and students \citep{loy2021bringing}. To answer this, there
have been numerous conventional hypothesis tests made available to test
for non-linear dependence \citep[e.g.][]{ramsey_tests_1969}, and
heteroskedasticity \citep[e.g.][]{breusch_simple_1979}.

Linear regression is a well-established procedure, and there is
considerable literature describing diagnostic procedures, e.g.
\citet{draper1998applied}, \citet{montgomery1982introduction},
\citet{belsley_regression_1980}, \citet{cook_applied_1999} and
\citet{cook1982residuals}. An interesting details, is that despite the
abundance of conventional tests, ALL of these writings advise that
plotting residuals is an essential tool for diagnosing regression model
problems. \citet{draper1998applied} and \citet{belsley_regression_1980}
explain that residual plots are usually revealing when the assumptions
are violated. \citet{cook_applied_1999} thinks formal tests and
graphical procedures are complementary and both have a place in residual
analysis, but they focus on graphical methods rather than on formal
testing. \citet{montgomery1982introduction} even suggests that residual
plots are more informative in most practical situations than the
corresponding conventional hypothesis tests.

The common wisdom of experts is that the optimal method for diagnosing
model fits is by plotting the data. The persistence of this advice to
check the plots is \emph{curious}, and investigating why this is, is the
subject of this paper.

@@

Graphical summaries where residuals are plotted against fitted values,
or other functions of the predictors (expected to be approximately
orthogonal to the residuals) are referred to as standard residual plots
in \citet{cook1982residuals}. Figure
\ref{fig:residual-plot-common-departures}A shows an example of an ideal
residual plot where points are symmetrically distributed around the
horizontal zero line (red), with no discernible patterns. There can be
various types of departures from this ideal pattern. Non-linearity,
heteroskedasticity and non-normality are three commonly checked
departures.

>>>>

Diagnosing a model is an important part of analysing data. In linear
regression analysis, studying the residuals from a model fit is a common
diagnostic activity. Residuals summarise what is not captured by the
model, and thus provide the capacity to identify what might be wrong.

We can assess residuals in multiple ways. To examine the their
univariate distribution, residuals may be plotted as a histogram or
normal probability plot. Using the classical normal linear regression
model as an example, if the distribution is symmetric and unimodal, we
would consider it to be well-behaved. However, if the distribution is
skewed, bimodal, multimodal, or contains outliers, there would be cause
for concern. We can also inspect the distribution by conducting a
goodness-of-fit test, such as the Shapiro-Wilk (SW) normality test
\citep{shapiro1965analysis}.

In addition, scatterplots of residuals plotted against the fitted values
and each of the explanatory variables, can be used to scrutinize their
relationships. If there are any visually discoverable patterns, the
model is potentially inadequate or incorrectly specified. In general,
one looks for noticeable departures from the model such as non-linear
pattern or heteroskedasticity. A non-linear pattern would suggest that
the model needs to have some additional non-linear terms.
Heteroskedasticity suggests that the error is dependent on the
predictors, and hence violates the independence assumption. However,
correctly judging whether NO pattern exists in a residual plot is a
difficult task for humans. Humans will almost always will see a pattern
(see \citet{kahneman}), so the question that really needs answering is
whether any pattern perceived is consistent with randomness, purely
sampling variability, or noise. It is especially difficult to teach this
to new analysts and students \citep{loy2021bringing}. To address this,
statistical tests are available to check for non-linear patterns
\citep[e.g.][]{ramsey_tests_1969}, and heteroskedasticity
\citep[e.g.][]{breusch_simple_1979}.

Linear regression is a well-established procedure, and there is
considerable literature describing diagnostic procedures, e.g.
\citet{draper1998applied}, \citet{montgomery1982introduction},
\citet{belsley_regression_1980}, \citet{cook_applied_1999} and
\citet{cook1982residuals}. Interestingly, despite the abundance of
conventional tests, ALL of these writings advise that plotting residuals
is an essential tool for diagnosing regression model problems:

\begin{quote}
\emph{Some most useful checks allowed by data plots should be done on a
routine basis for every regression.} \citep{draper1998applied}
\end{quote}

\begin{quote}
\emph{Such formal and informal procedures are complementary, and both
have a place in residual analysis.} \citep{cook1982residuals}
\end{quote}

\begin{quote}
\emph{In our experience, statistical tests on regression model residuals
are not widely used. In most practical situations the residual plots are
more informative than the corresponding tests. However, since residual
plots do require skill and experience to interpret, the statistical
tests may occasionally prove useful.} \citep{montgomery1982introduction}
\end{quote}

\noindent The common wisdom of experts is that plotting the residuals is
indispensable for diagnosing model fits. The ubiquity of this advice is
\emph{curious}: investigating why, is the subject of this paper.

@@

Graphical summaries where residuals are plotted against fitted values,
or other functions of the predictors (expected to be approximately
orthogonal to the residuals) are considered to be the most important
residual plots by \citet{cook_applied_1999}. Figure
\ref{fig:residual-plot-common-departures}A shows an example of an ideal
residual plot where points are symmetrically distributed around the
horizontal zero line (red), with no discernible patterns. There can be
various types of departures from this ideal pattern. Non-linearity,
heteroskedasticity and non-normality are three commonly checked
departures.
